---
title: "Uber Premium Booking Status Prediction"
author: "MGT256 Group 4"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
subtitle: "Multi-Class Classification Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


# Abstract
This analysis aims to predict Uber Premium Booking Status.

- Total samples: 20,795 rides
- Target variable: Booking Status with 2 categories (Completed and Non-Completed, 
including Cancelled by Driver, Cancelled by Customer, No Driver Found and Incomplete)
- Selected 7 core predictors: Booking Hour, Booking Month, Weekday, Vehicle Type, 
Traffic Condition and Booking Platform, Average Pickup Time.

Data Structure

- Source: Two datasets (booking_premium.csv, ride_premium.csv)
- Merge Key: Booking.ID
- Time Period: 2024-01-01 to 2024-12-30
- Features: 20+ variables including temporal, categorical, and numeric


# Data Import and Preparation

##  Load Required Libraries
```{r libraries}

# Clear environment
rm(list = ls())

# Data manipulation
library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
library(caret)    # For createDataPartition


# Visualization
library(ggplot2)
library(pROC)     # For ROC curve and AUC


```

## Import and Merge Data
```{r}

# Import datasets
booking_data <- read.csv('booking_premium.csv')
ride_data <- read.csv('ride_premium.csv')

# Merge datasets by Booking ID
merged_data <- merge(booking_data,ride_data,by='Booking.ID')

```

## Remove leading/trailing spaces from all character columns

```{r}

# Check if column is character 'TRUE'
char_cols <- sapply(merged_data, is.character)

# Clean only columns where 'char_cols = TRUE' (remove extra spaces)
merged_data[char_cols] <- lapply(merged_data[char_cols], function(x) {
  x <- str_trim(x)        # Remove leading/trailing spaces
  x <- str_squish(x)      # Replace multiple spaces with single space
  return(x)
  })

```

## Standardize missing values and negative values to NA

```{r}

# Define all representations of missing values
missing_values <- c("", " ", "null", "NA", "N/A", "na", "--", "nan", "NaN")

# Replace all missing value representations with NA
merged_data[] <- lapply(merged_data, function(column) {
  column[column %in% missing_values] <- NA
  return(column)
})

# Replace all negative values in Avg.VTAT with NA
merged_data <- merged_data %>%
    mutate(
        Avg.VTAT = ifelse(as.numeric(Avg.VTAT) < 0, NA, as.numeric(Avg.VTAT))
    )

```

## Handle target variable (Booking.Status)

```{r}

# Check Booking.Status
sum(is.na(merged_data$Booking.Status))
table(merged_data$Booking.Status)

# Create binary target:
merged_data$Target <- ifelse(merged_data$Booking.Status == "Completed", 
                             "Completed", "Not Completed")

# Ensure Class is a factor with "Good" first and "Bad" second.
# ------------------------------------------------------------
# In logistic regression (glm with family = binomial),
# predicted probabilities (type = "response") correspond to
# the *second* level of the outcome factor.
#
# By setting the levels as c("Good", "Bad"):
#   - The model will output predicted probabilities for the "Bad" class,
#     which is the event of interest (the important/risk class).
#   - These probabilities can be directly interpreted as P(Bad | predictors).
#   - Sensitivity and specificity can then be computed with respect to
#     correctly identifying "Bad" credit applicants.
#
# If the order were reversed, the model would instead return P(Good),
# which would flip the interpretation of all predicted probabilities.

# Convert to factor (Completed first, Not Completed second)
merged_data$Target <- factor(merged_data$Target, 
                             levels = c("Completed", "Not Completed"))
levels(merged_data$Target)


# Target distribution
target_counts <- table(merged_data$Target)
target_prop <- prop.table(target_counts) * 100

# Plot distribution
barplot(target_counts,
        main = "Booking Status Distribution",
        col = c("lightblue", "coral"),
        ylab = "Number of Bookings",
        cex.main = 1.8,
        cex.axis = 1.3,
        cex.names = 1.3,
        ylim = c(0, max(target_counts) * 1.15))
```

## Standardize Date format and extract features

```{r}

# Function to parse 3 date formats found in the data

parse_date <- function(date_string) {
    formats <- c(
        "%d-%b-%Y",      # 03-Mar-2024
        "%Y-%m-%d",      # 2024-01-15
        "%m/%d/%Y"       # 12/15/2024
        )
    
    result <- NA         # in failed cases
    for(fmt in formats) {
        parsed <- suppressWarnings(as.Date(date_string, format = fmt))
        if(!is.na(parsed)) {
            result <- parsed
            break # Stop the loop once we find a format that works
        }
        }
    return(result)
    }

# Apply the date parsing function
merged_data$Date_Parsed <- sapply(merged_data$Date, parse_date) %>%
    as.Date(origin = "1970-01-01")

# Extract temporal features
merged_data <- merged_data %>%
    mutate(
        Weekday = wday(Date_Parsed, label = TRUE),
        Month = month(Date_Parsed)
    )
```

## Standardize Time format and extract Hour

```{r}

# Function to parse 3 time formats found in the data

parse_time <- function(time_string) {
    formats <- c(
        "%H:%M:%S",      # 13:15:17
        "%H:%M",         # 21:15
        "%I:%M %p"       # 2:34 AM, 10:22 PM
        )
    result <- NA         # in failed cases
    for(fmt in formats) {
        parsed <- suppressWarnings(strptime(time_string, format = fmt))
        if(!is.na(parsed)) {
            result <- (as.numeric(format(parsed, "%H"))) # Extract hour (0-23)
            break
        }
        }
    return(result)
    }

# Apply the function to extract hour from Time column
merged_data$Hour <- sapply(merged_data$Time, parse_time)

```

## Check NA counts in predictors

```{r}

sum(is.na(merged_data$Hour))
sum(is.na(merged_data$Weekday))
sum(is.na(merged_data$Month))
sum(is.na(merged_data$Vehicle.Type))
sum(is.na(merged_data$Booking.Platform))
sum(is.na(merged_data$Traffic.Level))
sum(is.na(merged_data$Avg.VTAT))

```

## Remove rows with NA in Traffic.Level and Avg.VTAT

```{r}

merged_data <- merged_data[!is.na(merged_data$Traffic.Level), ]
merged_data <- merged_data[!is.na(merged_data$Avg.VTAT), ]

```

## Convert Categorical Variables to Factors

```{r}

# List of categorical predictors
categorical <- c("Vehicle.Type", "Traffic.Level", "Booking.Platform", "Weekday")

# Convert to factors
merged_data[categorical] <- sapply(merged_data[categorical], as.factor)
```

# Data Visualizations

## Graph 1: Cancellation Rate by Hour
```{r}
hourly <- merged_data %>%
  group_by(Hour) %>%
  summarise(Cancel_Rate = mean(Target == "Not Completed") * 100)

plot(hourly$Hour, hourly$Cancel_Rate,
     type = "b",
     col = "steelblue",
     lwd = 3,
     pch = 19,
     cex = 1.5,
     main = "Cancellation Rate by Hour",
     xlab = "Hour (0-23)",
     ylab = "Cancellation Rate (%)",
     cex.main = 1.8,
     cex.axis = 1.3,
     cex.lab = 1.3,
     ylim = c(25, 45))
grid()
```

## Graph 2: Avg.VTAT Distribution
```{r}

hist(merged_data$Avg.VTAT,
     breaks = 30,
     col = "skyblue",
     border = "white",
     main = "Distribution of Average Pickup Time",
     xlab = "Avg.VTAT (minutes)",
     ylab = "Frequency",
     cex.main = 1.8,
     cex.axis = 1.3,
     cex.lab = 1.3)

abline(v = mean(merged_data$Avg.VTAT), col = "red", lwd = 3, lty = 2)
```

## Graph 3: Avg.VTAT by Booking Status
```{r}

boxplot(Avg.VTAT ~ Target, 
        data = merged_data,
        col = c("lightblue", "coral"),
        main = "Avg Pickup Time by Booking Status",
        xlab = "Status",
        ylab = "Avg.VTAT (minutes)",
        cex.main = 1.8,
        cex.axis = 1.3,
        cex.lab = 1.3)
```

## Graph 4: Booking Volume by Hour
```{r}
hour_count <- table(merged_data$Hour)

barplot(hour_count,
        col = "steelblue",
        main = "Booking Volume by Hour",
        xlab = "Hour (0-23)",
        ylab = "Number of Bookings",
        cex.main = 1.8,
        cex.axis = 1.3,
        cex.lab = 1.3)
```


# Logistic Regression

## Stratified 60/40 Train-Validation Split

```{r}

set.seed(19)

idx <- createDataPartition(merged_data$Target, p = 0.6, list = FALSE)
train_data <- merged_data[idx, ]
valid_data <- merged_data[-idx, ]

print(prop.table(table(train_data$Target))*100)

print(prop.table(table(valid_data$Target))*100)

```

## Fit Logistic Regression Models on training

```{r}

# Model 1: Original 
model1 <- glm(Target ~ Hour + Weekday + Month + Vehicle.Type + 
              Traffic.Level + Booking.Platform + Avg.VTAT,
              data = train_data, family = binomial)

# Model 2: Add polynomial terms
model2 <- glm(Target ~ Hour + I(Hour^2) + Weekday + Month + 
              Vehicle.Type + Traffic.Level + Booking.Platform + 
              Avg.VTAT + I(Avg.VTAT^2),
              data = train_data, family = binomial)

# Model 3: Add interaction
model3 <- glm(Target ~ Hour * Avg.VTAT + Weekday + Month + 
              Vehicle.Type + Traffic.Level + Booking.Platform,
              data = train_data, family = binomial)

# Model 4: Log transform
model4 <- glm(Target ~ Hour + Weekday + Month + 
              Vehicle.Type + Traffic.Level + Booking.Platform + 
              log(Avg.VTAT),
              data = train_data, family = binomial)
```


## Predicted probability of "Not Completed" on validation

```{r}

# Predictions
valid_prob1 <- predict(model1, newdata = valid_data, type = "response")
valid_prob2 <- predict(model2, newdata = valid_data, type = "response")
valid_prob3 <- predict(model3, newdata = valid_data, type = "response")
valid_prob4 <- predict(model4, newdata = valid_data, type = "response")

head(valid_prob1, 10)
head(valid_prob2, 10)
head(valid_prob3, 10)
head(valid_prob4, 10)

# Calculate AUC
auc1 <- auc(roc(valid_data$Target, valid_prob1))
auc2 <- auc(roc(valid_data$Target, valid_prob2))
auc3 <- auc(roc(valid_data$Target, valid_prob3))
auc4 <- auc(roc(valid_data$Target, valid_prob4))

# Compare
comparison <- data.frame(
  Model = c("Original", "Polynomial", "Interaction", "Log Transform"),
  AUC = c(auc1, auc2, auc3, auc4)
)

print(comparison)

```

## Find Optimal Threshold (Log Transform Model)

```{r}

# Compute the ROC curve
roc_obj <- roc(response = valid_data$Target, predictor = valid_prob4)


# Optimal threshold
optimal_thres <- coords(roc_obj, x = "best",
                      best.method = "closest.topleft",
                      ret = "threshold")
optimal_thres

# Plot the ROC curve with the AUC displayed
plot.roc(roc_obj, 
         legacy.axes = TRUE, 
         print.auc = TRUE,
         main = "ROC — Log Transform Model", 
         lwd = 3,
         col = 'steelblue')
```

## Confusion Matrix at ROC-Chosen Threshold

```{r}
# Classify using optimal threshold
pred_roc <- factor(ifelse(valid_prob4 > as.numeric(optimal_thres), 
                          "Not Completed", "Completed"),
                   levels = c("Completed", "Not Completed"))

# Confusion matrix
cm_roc <- confusionMatrix(pred_roc, 
                          valid_data$Target, 
                          positive = "Not Completed")
print(cm_roc)
```

```{r}
# ==========================================
# KNN with Manual 10-Fold Cross-Validation
# ==========================================

library(class)   # for knn()
library(caret)   # for createFolds() and confusionMatrix()

# 1) Build KNN datasets from existing train_data / valid_data
knn_train <- train_data %>%
  dplyr::select(Target, Hour, Weekday, Month,
                Vehicle.Type, Traffic.Level,
                Booking.Platform, Avg.VTAT) %>%
  na.omit()

knn_valid <- valid_data %>%
  dplyr::select(Target, Hour, Weekday, Month,
                Vehicle.Type, Traffic.Level,
                Booking.Platform, Avg.VTAT) %>%
  na.omit()

# Make sure Target is a factor with same levels
knn_train$Target <- factor(knn_train$Target)
knn_valid$Target <- factor(knn_valid$Target,
                           levels = levels(knn_train$Target))

# 2) Turn predictors into numeric matrix (dummy variables for factors)
x_train <- model.matrix(
  Target ~ Hour + Weekday + Month + Vehicle.Type +
    Traffic.Level + Booking.Platform + Avg.VTAT,
  data = knn_train
)[, -1]  # drop intercept

x_valid <- model.matrix(
  Target ~ Hour + Weekday + Month + Vehicle.Type +
    Traffic.Level + Booking.Platform + Avg.VTAT,
  data = knn_valid
)[, -1]

y_train <- knn_train$Target
y_valid <- knn_valid$Target

# 3) Scale predictors (very important for KNN)
x_train_scaled <- scale(x_train)
x_valid_scaled <- scale(
  x_valid,
  center = attr(x_train_scaled, "scaled:center"),
  scale  = attr(x_train_scaled, "scaled:scale")
)

# 4) 10-fold cross-validation to choose k
set.seed(19)
folds <- createFolds(y_train, k = 10, list = TRUE, returnTrain = FALSE)

k_values   <- seq(3, 25, 2) # try odd k's: 3,5,7,...,25
cv_results <- data.frame(k = k_values, Accuracy = NA_real_)

for (i in seq_along(k_values)) {
  k <- k_values[i]
  acc_vec <- numeric(length(folds))
  
  for (f in seq_along(folds)) {
    val_idx <- folds[[f]]
    
    x_tr <- x_train_scaled[-val_idx, , drop = FALSE]
    y_tr <- y_train[-val_idx]
    x_te <- x_train_scaled[val_idx, , drop = FALSE]
    y_te <- y_train[val_idx]
    
    pred <- knn(
      train = x_tr,
      test  = x_te,
      cl    = y_tr,
      k     = k
    )
    
    acc_vec[f] <- mean(pred == y_te)
  }
  
  cv_results$Accuracy[i] <- mean(acc_vec)
}

cv_results

# 5) Pick the best k
best_k <- cv_results$k[which.max(cv_results$Accuracy)]
best_k

# 6) Train final KNN on full training data and evaluate on validation set
knn_valid_pred <- knn(
  train = x_train_scaled,
  test  = x_valid_scaled,
  cl    = y_train,
  k     = best_k
)

# Confusion matrix on validation set
knn_cm <- confusionMatrix(
  knn_valid_pred,
  y_valid,
  positive = "Not Completed"
)
knn_cm
```


# Results and Discussion

## Model Performance

Using the log-transformed model with ROC-optimal threshold:

- **AUC:** 0.5080757 (marginally above random guessing at 0.50)
- **Accuracy:** 0.4884
- **Sensitivity:** 0.5047 (detects ~50% of cancellations)
- **Specificity:** 0.4804
- **Precision:** 32.2% (only ~32% of predictions are correct)

## Key Findings

1. **Weak Predictive Power:** All models achieved AUC ~ 0.50-0.51, indicating 
performance barely better than random guessing.

2. **Model Transformations:** Polynomial terms, interactions, and 
log transformation provided negligible improvements (+0.003 AUC), confirming the 
fundamental limitation is weak features, not model specification.

3. **Class Imbalance:** The 63%/37% split required threshold optimization, 
but even with optimal thresholds, the model remains weak.

## Limitations

- Current features lack strong predictive signals
- Missing key historical data: customer booking history, driver-side predictors,
trip distance
- Logistic regression may not capture complex patterns

## Improvement

1. **Feature Engineering:**
   - Add customer-level features (past cancellation rate, booking frequency)
   - Incorporate driver-side variables (acceptance rate, distance to pickup)

2. **Advanced Methods:**
   - Explore advanced models to capture relationships

3. **Data Collection:**
   - Enhance dataset with richer behavioral and contextual features
   - Track customer engagement metrics and historical patterns